Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-25 submission/manuscript.Rmd
abstract draft

>>>>

It is currently unclear why mutations in numerous chromatin-modifying enzymes cause neurodevelopmental disorders (NDDs). While some chromatin regulators are important for neuron-specific gene expression, loss of repressive chromatin regulators can also lead to the aberrant transcription of tissue-specific genes outside of their intended context. Because very few studies have characterized the these aberrant, tissue-specific transcripts, the mechanisms behind their dysregulation and their functional consequences are largely unknown. Here, we explored this crisis in cellular identity in NDDs by investigating the ectopic expression of germline (e.g. sperm and egg) genes in mice lacking lysine demethylase 5c (KDM5C, also known as JARID1C or SMCX), an eraser of histone 3 lysine 4 di and tri-methylation (H3K4me2/3). Male _Kdm5c_ knockout (-KO) mice recapitulate key phenotypes of Claes-Jensen X-linked intellectual disability (MRXSCJ, OMIM: 300534), including heightened aggression and learning impairments. We found male _Kdm5c_ knockout (-KO) mice that recapitulate key phenotypes of Claes-Jensen X-linked intellectual disability (MRXSCJ, OMIM: 300534) aberrantly express liver, muscle, ovary, and testis genes within the amygdala and hippocampus. The majority of tissue-enriched genes were typically unique to germ cells, indicating the erosion of the soma-germline boundary is a major effect of KDM5C loss. Using RNA sequencing datasets of germ cell-depleted mice, we curated a list of germline-enriched genes and found KDM5C regulates many key drivers of meiosis and germline identity in an _in vitro_ model of the post-implantation epiblast. While KDM5C promotes the initial placement of DNA methylation at a subset of germline gene promoters during early embryogenesis, germline genes can also be aberrantly expressed in _Kdm5c_-KO cells indirectly of KDM5C binding. Intriguingly, many of the germline genes expressed in the _Kdm5c_-KO brain are late-stage spermatogenesis genes that are not directly bound by KDM5C. This suggests germline developmental programs can occur ectopically in the background of typical _Kdm5c_KO development. Ultimately, this work provides insight into the demarcation of somatic and germline identity in mammals while also providing the foundational mechanisms by which to test the impact of germline gene dysregulation in neurodevelopmental disorders.
